Supplementary Information

### 2021-02-09 doc/manuscript/MEE_manuscript.Rmd
working on site descriptions

We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or within 10 km (n=6), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).
In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017] and numerous other studies [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019], albeit contrasting with some others [e.g., @voelker_historical_2006; @hember_tree_2019].
With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be a dominant driver for the trees in natural forest settings analyzed here. 
Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and the preponderance of tree-ring evidence points to little CO~2~ simulation of tree growth rates [@walker_integrating_2020].

>>>>

We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).
In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017] 
With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @voelker_historical_2006; @hember_tree_2019].
A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
working on site descriptions

We note that the secondary forest status of much of BCNM differs from ForestGEO's 50-ha plot on Barro Colorado Island (within BCNM), which is old-growth, and shows little directional trend in tree growth [@rutishauser_testing_2020].

@@

*Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions [@baker_disturbance_2005].

The site underwent a widespread, catastrophic disturbance in the mid-1800s and several smaller, more localized disutrbances in the 1910s, 1940s, and 1960s [@baker_disturbance_2005]

@@

@@

@@

### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic
The Zofin forest is not fully without past human direct interventions, a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].

### Niobrara, Nebraska, USA
Riparian site
BEPA are growing on N facing slopes near stream

### Little Tesuque, New Mexico, USA
[@williams_temperature_2013]

@@

\newpage
## `r site_details_legend` 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/site_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,   escape = F)%>%
  column_spec(2, width = "3cm") %>%
  column_spec(5, width = "2cm") %>%
  column_spec(6, width = "2cm") %>%
  column_spec(7, width = "1.5cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
*Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).

>>>>

Cores were collected within the Barro Colorado Nature Monument (BCNM), a 5600-ha reserve established in 1923 that includes Barro Colorado Island (BCI) and the surrounding mainland peninsulas [@alfaro-sanchez_growth_2017]. 
BCNM contains a mix of old-growth and mature secondary tropical moist forest. 
Live canopy trees on the peninsulas were cored between January 2015 and April 2015. 
In addition, several fallen dead trees on BCI were sampled by taking discs with chain saws.

We note that the secondary forest status of much of BCNM differs from ForestGEO's 50-ha plot on BCI, which is old-growth, and shows little directional trend in woody productivity [@rutishauser_testing_2020].

@@

Cores were collected within a ~300 ha acre area of seasonal dry evergreen and mixed deciduous forest in the Huai Kha Khaeng Wildlife Sanctuary (HKK).
The site underwent a widespread, catastrophic disturbance in the mid-1800s and several smaller, more localized disturbances in the 1910s, 1940s, and 1960s [@baker_disturbance_2005].

Cores were collected between December 2010 and December 2011, sampling trees >5 cm DBH with the goal or representing a random subset of the local population of the target species. 
Details are given in @vlam_temperature_2014.

@@

Cores were collected within the ForestGEO plot at the Smithsonian Conservation Biology Institute (SCBI). 
The forest is a secondary temperate broadleaf deciduous forest, which developed following agricultural abandonment in the mid 19th-century, dominated by the species included in our analysis [@bourg_initial_2013]. 
Specifically, the 14 species sampled together account for 97% of woody productivity at the site [@helcoski_growing_2019].

In 2010-2011, cores were collected from a random subset of live trees $\ge$ 10 cm DBH.
In 2016 and 2017, cores were extracted from trees found newly dead during an annual tree mortality census [@gonzalez-akre_patterns_2016].
Details are given in @helcoski_growing_2019.

@@

Lilly Dickey Woods (LDW) is a secondary temperate broadleaf deciduous forest dominated by oaks (*Quercus* spp.) and hickories (*Carya* spp.).

Cores were collected outside the ForestGEO plot, but within a contiguous patch of forest.
In 2012 - 2014, cores were collected from live or dead dominant canopy trees with morphological characteristics typical of older trees [@maxwell_declining_2016].
In 2020, additional cores were collected following an ecological study design, wherein all live trees $\ge$ **##** cm within a **(SIZE)** plot were sampled.

@@

Harvard Forest (HF) is a secondary temperate mixed forest dominated by *Tsuga canadensis* and northern hardwood species including *Quercus rubra*, *Acer rubrum*, and *Betula alleghaniensis*.
Cores were collected within the ForestGEO megaplot, and in the footprint of the original HF eddy-flux tower [@finzi_carbon_2020]. 
Samples were collected following an ecological study design [@dye_comparing_2016], wherein all living or dead trees $\ge$ 10 cm DBH within a **(SIZE)** plot were cored.

### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic

The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., *Picea abies* L. and *Abies alba* Mill.
The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].

In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
The tree-ring data were collected for several studies around 2010 [@samonil_individual-based_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_species-specific_nodate]. 
The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
Therefore, these data consist mainly of dominant and codominant tree species and relatively low number of young trees.

### Niobrara, Nebraska, USA

The Niobrara site sits at the prairie-forest ecotone of the North American Great Plains. 
There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019.
The largest healthy trees within each stand were selected for coring.

### Little Tesuque, New Mexico, USA
The Little Tesuque (LT) site is in a mature dry mixed conifer forest outside of Santa Fe, NM, in the foothills of the Sangre de Cristo Mountains.
The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years.
The growth of trees in the region, including at this site, are being severely affected by a strong drying trend under climate change [@touchan_millennial_2011; @williams_temperature_2013], with many missing growth rings in recent years.

In 2020, live canopy and sub-canopy trees, all >20 cm DBH, were sampled within a 1 ha plot (Little Tesuque plot 3; *lts3*).

@@

\newpage
## `r site_details_legend` 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/site_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,   escape = F)%>%
  column_spec(2, width = "3cm") %>%
  column_spec(5, width = "2cm") %>%
  column_spec(6, width = "2cm") %>%
  column_spec(7, width = "1.5cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
*Refers to ForestGEO plot (exception: LT) and is not necessarily the most accurate value for exact locations at which cores were taken. Geographic coordinates were used to extract climate data from CRU. 
**Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
